Grove: A Convergent Collaborative Structure-Editor Calculus

### 2021-03-01 grove-paper.tex
rename W to U and wreath to unicycular graph

A grove is a 4-tuple of the following form:
\[
  (e_r, MP, NP, W)
\]
where
\begin{align*}
  e_r &= \expr(v_0) \\
  MP &= \{\expr(v) : |\parents(v)| > 1\} \\
  NP &= \{\expr(v) : |\parents(v)| = 0 \land v \ne v_0\} \\
  W  &= \{\expr(v) : |\parents(v)| = 1 \wedge v = \min(\ancestors(v))\}
\end{align*}
For example, the grove corresponding to the graph in \autoref{fig:Decomposition example graph} is
given in \autoref{fig:Decomposition example grove}.

Let us now examine each of the four components in more detail.
The first component $e_r$ is the expression derived from the root vertex $v_0$.
\autoref{fig:Decomposition expr} defines $\expr(v)$ wich determines the expression derived from $v$.
There is one rule for each constructor form.
For example, the expression derived from ${+}_{\id{38}} = (38, \ExpPlus)$ is $\expr'({+}_{\id{38}}, \PlusLeft) \text{+}^{\id{38}} \expr'({+}_{\id{38}}, \PlusRight)$.
In each child position, $\expr'(v, p)$ determines the sub-expression derived from the children of $v$ at position $p$.
An empty hole $\hole$ appears if there are no children at the given position.
A conflict hole $\conflictHole{e_1,\ldots,e_n}$ appears if there are multiple children.
If there is one child, the expression is determined recursively unless that child appears in $MP$ or $W$, discussed below, in which case we leave a reference, $\multiVertex{u}$ or $\cycleVertex{u}$ respectively.
For the $\ExpLam$ case, we define $\patt(q)$ and $\type(\tau)$ similarly.

@@

The fourth component $W$ is the set of expressions derived from the remaining vertices, which form cyclic structures we call wreaths, where each vertex has exactly one parent.
Every wreath consists of a single cycle where each vertex in the cycle may have additional children that are not part of the cycle.
We need to break the cycle to be able to turn a wreath into a tree.
To break the cycle, we arbitrarily choose the vertex on the cycle with the smallest id according to the total ordering on $\U$.
Formally, we choose the vertex $v$ such that $v = \min(\ancestors(v))$ because
every vertex on the cycle is its own ancestor.

@@

% TODO: add text around these definitions -- turn them into paragraphs.

@@

Finally, once we've already accounted for things with more or less than one
parent, everything else must be single parented. We call these things wreaths. A
wreath is a cycle of single-parented vertices that each may have single-parented
descendants.

lemma: a wreath is a cycle with trees hanging off of it

lemma: wreath traversal always produces a valid tree

Wreath traversal occurs in two directions. A vertex is chosen as an arbitrary
starting point and its parents are traversed and marked as seen. Any vertex that
has been seen twice must be on the cycle. We take the first one as the root of
the wreath and traverse its children to produce a tree that covers the wreath.
Any remaining single-parented vertices must be parts of other wreaths and are
wreath-traversed until there are no more single-parented vertices left.

>>>>

A grove is a 4-tuple of the following form:
\[
  (e_r, MP, NP, W)
\]
where
\begin{align*}
  e_r &= \expr(v_0) \\
  MP &= \{\expr(v) : |\parents(v)| > 1\} \\
  NP &= \{\expr(v) : |\parents(v)| = 0 \land v \ne v_0\} \\
  U  &= \{\expr(v) : |\parents(v)| = 1 \wedge v = \min(\ancestors(v))\}
\end{align*}
For example, the grove corresponding to the graph in \autoref{fig:Decomposition example graph} is
given in \autoref{fig:Decomposition example grove}.

Let us now examine each of the four components in more detail.
The first component $e_r$ is the expression derived from the root vertex $v_0$.
\autoref{fig:Decomposition expr} defines $\expr(v)$ wich determines the expression derived from $v$.
There is one rule for each constructor form.
For example, the expression derived from ${+}_{\id{38}} = (38, \ExpPlus)$ is $\expr'({+}_{\id{38}}, \PlusLeft) \text{+}^{\id{38}} \expr'({+}_{\id{38}}, \PlusRight)$.
In each child position, $\expr'(v, p)$ determines the sub-expression derived from the children of $v$ at position $p$.
An empty hole $\hole$ appears if there are no children at the given position.
A conflict hole $\conflictHole{e_1,\ldots,e_n}$ appears if there are multiple children.
If there is one child, the expression is determined recursively unless that child appears in $MP$ or $U$, discussed below, in which case we leave a reference, $\multiVertex{u}$ or $\cycleVertex{u}$ respectively.
For the $\ExpLam$ case, we define $\patt(q)$ and $\type(\tau)$ similarly.

@@

The fourth component $U$ is the set of expressions derived from the remaining vertices, which form unicycular graphs where each vertex has exactly one parent.
Every unicycular graph consists of a single cycle where each vertex in the cycle may have additional children that are not part of the cycle.
Breaking the cycle turns a unicyclic graph into a tree.
To break the cycle, we arbitrarily choose the vertex on the cycle with the smallest id according to the total ordering on $\U$.
Formally, we choose the vertex $v$ such that $v = \min(\ancestors(v))$ because
only a vertex on the cycle can be its own ancestor.

@@

@@

Finally, once we've already accounted for things with more or less than one
parent, everything else must be single parented. We call these things unicycular graphs. A
unicycular graph is a cycle of single-parented vertices that each may have single-parented
descendants.

lemma: a unicycular graph is a cycle with trees hanging off of it

lemma: unicycular graph traversal always produces a valid tree

Unicycular graph traversal occurs in two directions. A vertex is chosen as an arbitrary
starting point and its parents are traversed and marked as seen. Any vertex that
has been seen twice must be on the cycle. We take the first one as the root of
the unicycular graph and traverse its children to produce a tree that covers the unicycular graph.
Any remaining single-parented vertices must be parts of other unicycular graphs and are
traversed until there are no more single-parented vertices left.
